Leadership across cultures and contexts: an exploratory analysis

### 2020-06-18 leadership_across_cultures_contexts.Rmd
Added Buss 2020 citation.
Moved some ethnographic sample text to SI (e.g., the words, paragraph length, etc.).

<!-- WEIRD PEOPLE PROBLEMS -->
There is now a broad consensus on the importance of data from the full range of human cultural diversity. Since the seminal WEIRD people paper [@henrich_weirdest_2010, *The weirdest people in the world?*], many social scientists have taken up the challenge to obtain data from "non-WEIRD" populations. Dichotomizing populations and their psychologies and social dynamics as WEIRD or non-WEIRD, however, is a mistake.^[A Google Scholar search for "non-WEIRD" produced over 1000 results.] As anthropology has conclusively demonstrated, there is enormous diversity across societies that does not remotely resemble a dichotomy.

@@

There is potentially more diversity within non-WEIRD populations as there is within WEIRD ones, and there is also substantial overlap of the behaviors of people everywhere [@brown_human_1991]. @henrich_weirdest_2010 in their critique of the over-reliance on WEIRD samples emphasized a holistic approach to cultural diversity within and across "telescoping" levels of human societies. It is therefore important to assess cultural variation without assuming that everyone within a culture is essentially the same, and without putting undue emphasis on either the between-population differences or similarities.

@@

<!-- MORE UNIVERSAL QUALITIES FROM WEIRD AND NON WEIRD -->
Intelligent and knowledgeable individuals [e.g., @johnson_leadership_2015; @judge_intelligence_2004; @neel_being_1980; @antonakis_does_2009], elders, and respected individuals tend to have greater influence over community decisions across cultures, including among egalitarian hunter-gatherers, stratified chiefdoms, and nation state organizations [@silverman_how_1978; @tooby_cognitive_2006; @garfield_review_2019; @bass_handbook_1990; @garfield_evolutionary_2019; @garfield_investigating_2020]. Leaders also tend to have large social networks [@von_rueden_leadership_2014; @von_rueden_roots_2014; @walker_living_2012], and embody the qualities most respected by the group [@henrich_evolution_2001; @van_vugt_evolutionary_2006; @collier_politics_1981; @low_men_1992; @roscoe_social_2019]. Leaders across multiple contexts tend to be taller and more physically formidable, which might facilitate resolving inter- and intra-group  conflicts [@lukaszewski_role_2016; @von_rueden_leadership_2014; @von_rueden_roots_2014]. Charisma and oratory skills are common properties of leaders when they must rely on persuasion to influence community members [@grabo_charismatic_2016].

<!-- STATUS VS LEADERSHIP -->
Leaders have disproportionate group influence, whereas status is often conceptualized as a relative indicator of the social value of the individual by the group and of their relative access to contested resources [e.g., @blader_whats_2014; @von_rueden_roots_2014]. Leaders are expected to be high status [e.g., @von_rueden_leadership_2015; @van_vugt_genes_2020; @cheng_dominance_2019; @cheng_psychology_2014]. The causal relationship between the two, however, is likely to be complex, context-specific, and driven by underlying correlates [e.g., @blader_whats_2014; @cheng_dominance_2019; @kantner_identifying_2010; @von_rueden_roots_2014].

<!-- MALE BIAS -->
Substantial evidence suggests a near universal male-bias in community leadership across human societies [@low_sex_1992; @whyte_status_1978; @sanday_female_1981], as well as across diverse nonhuman social species [@smith_obstacles_2020; @tiger_female_1978]. Sex, however, is confounded with factors such as social network size, economic specialization, and education, that are also associated with leadership. After controlling for these factors, some studies have found that sex is not a strong predictor of community leadership [@yanca_female_2004; @garfield_investigating_2020; @von_rueden_sex_2018].

@@

ZG and KS then discussed every instance of disagreement between their respective coding on all leadership dimensions and group structure variables to produce a consensus matrix, which is used in the analyses here and added to the published data repository (in a data frame named *leader_text2*). For an illustration of the coding process, see the *Coding example* section of the SI.

@@

Our fourth goal was to explore lower dimensional representations of these data that might reveal informative "features." These included logistic PCA, minimal spanning trees and nearest neighbors (mst-kNN), and hierarchical cluster analysis. See the supplementary information (SI) for brief descriptions of these methods. Our fifth goal was to assess variation in our lower-dimensional "features" by group context, subsistence strategy, continental region, and leader sex, using logistic mixed effects models fit with the lme4 package. Our sixth goal was to evaluate the features in light of influential evolutionary theories of leadership. Drawing on the empirically derived features as a guide, we aimed to construct "elements" of leadership that corresponded to elements of influential theories, examining intersections of these elements across all text records.

@@

The geographic distribution of the cultures in our sample is displayed in Figure \@ref(fig:geographic) (see also Figure S\@ref(fig:subsisttree)). The database of text record comprised `r nrow(all_data)` text records from `r length(unique(leadershipdata::documents$d_ID))` unique documents describing `r length(unique(all_data$d_culture))` cultures. The number of text records per culture ranged from `r min(leader_cult$number_leader_records, na.rm=T)` to `r max(leader_cult$number_leader_records, na.rm=T)`, with a median of `r median(leader_cult$number_leader_records, na.rm=T)`, a mean of `r signif(mean(leader_cult$number_leader_records, na.rm=T), 3)`, and a standard deviation of `r signif(sd(leader_cult$number_leader_records, na.rm=T), 3)`. The text records were generally short, ranging from `r textstats$min` to `r textstats$max` words, with a median word count of `r textstats$median`, a mean of `r textstats$mean`, and standard deviation of `r textstats$sd`. Of the `r nrow(all_data)` text records, `r final_record_count` had evidence for at least one dimension in the study.

@@

>>>>

<!-- WEIRD PEOPLE PROBLEMS -->
There is now a broad consensus on the importance of data from the full range of human cultural diversity. Since the seminal WEIRD people paper [@henrich_weirdest_2010], many social scientists have taken up the challenge to obtain data from "non-WEIRD" populations. Dichotomizing populations and their psychologies and social dynamics as WEIRD or non-WEIRD, however, is a mistake.^[A Google Scholar search for "non-WEIRD" produced over 1000 results.] As anthropology has conclusively demonstrated, there is enormous diversity across societies that does not remotely resemble a dichotomy.

@@

There is potentially more diversity within "non-WEIRD" populations as there is within WEIRD ones, and there is also substantial overlap of the behaviors of people everywhere [@brown_human_1991]. @henrich_weirdest_2010 in their critique of the over-reliance on WEIRD samples emphasized a holistic approach to cultural diversity within and across "telescoping" levels of human societies. It is therefore important to assess cultural variation without assuming that everyone within a culture is essentially the same, and without putting undue emphasis on either the between-population differences or similarities.

@@

<!-- MORE UNIVERSAL QUALITIES FROM WEIRD AND NON WEIRD -->
Intelligent and knowledgeable individuals [e.g., @johnson_leadership_2015; @judge_intelligence_2004; @neel_being_1980; @antonakis_does_2009], elders, and respected individuals tend to have greater influence over community decisions across cultures, including among egalitarian hunter-gatherers, stratified chiefdoms, and nation-state organizations [@silverman_how_1978; @tooby_cognitive_2006; @garfield_review_2019; @bass_handbook_1990; @garfield_evolutionary_2019; @garfield_investigating_2020]. Leaders also tend to have large social networks [@von_rueden_leadership_2014; @von_rueden_roots_2014; @walker_living_2012], and embody the qualities most respected by the group [@henrich_evolution_2001; @van_vugt_evolutionary_2006; @collier_politics_1981; @low_men_1992; @roscoe_social_2019]. Leaders across multiple contexts tend to be taller and more physically formidable, which might facilitate resolving inter- and intra-group  conflicts [@lukaszewski_role_2016; @von_rueden_leadership_2014; @von_rueden_roots_2014]. Charisma and oratory skills are common properties of leaders when they must rely on persuasion to influence community members [@grabo_charismatic_2016].

<!-- STATUS VS LEADERSHIP -->
Leaders have disproportionate group influence, whereas status is often conceptualized as a relative indicator of the social value of the individual by the group and of their relative access to contested resources [e.g., @blader_whats_2014; @von_rueden_roots_2014]. Leaders are expected to be high status [e.g., @von_rueden_leadership_2015; @van_vugt_genes_2020; @cheng_dominance_2019; @cheng_psychology_2014; @buss_human_2020]. The causal relationship between the two, however, is likely to be complex, context-specific, and driven by underlying correlates [e.g., @blader_whats_2014; @cheng_dominance_2019; @kantner_identifying_2010; @von_rueden_roots_2014].

<!-- MALE BIAS -->
Substantial evidence suggests a near universal male-bias in community leadership across cultures [@low_sex_1992; @whyte_status_1978; @sanday_female_1981], as well as across diverse nonhuman social species [@smith_obstacles_2020; @tiger_female_1978]. Sex, however, is confounded with factors such as social network size, economic specialization, and education, that are also associated with leadership. After controlling for these factors, some studies have found that sex is not a strong predictor of community leadership [@yanca_female_2004; @garfield_investigating_2020; @von_rueden_sex_2018].

@@

ZG and KS then discussed every instance of disagreement between their respective coding on all leadership dimensions and group structure variables to produce a consensus matrix, which is used in the analyses here and added to the published data repository (in a data frame named *leader_text2*). For an illustration of the coding process, see the *Coding example* section of the supplementary information (SI).

@@

Our fourth goal was to explore lower dimensional representations of these data that might reveal informative "features." These included logistic PCA, minimal spanning trees and nearest neighbors (mst-kNN), and hierarchical cluster analysis. See the SI for descriptions of these methods. Our fifth goal was to assess variation in our lower-dimensional "features" by group context, subsistence strategy, continental region, and leader sex, using logistic mixed effects models fit with the lme4 package. Our sixth goal was to evaluate the features in light of influential evolutionary theories of leadership. Drawing on the empirically derived features as a guide, we aimed to construct "elements" of leadership that corresponded to elements of influential theories, examining intersections of these elements across all text records.

@@

The geographic distribution of the cultures in our sample is displayed in Figure \@ref(fig:geographic) (see also Figure S\@ref(fig:subsisttree)). The database of text record comprised `r nrow(all_data)` text records from `r length(unique(leadershipdata::documents$d_ID))` unique documents describing `r length(unique(all_data$d_culture))` cultures.

@@

In this data set (originally constructred by @garfield_evolutionary_2019) the number of text records per culture ranged from `r min(leader_cult$number_leader_records, na.rm=T)` to `r max(leader_cult$number_leader_records, na.rm=T)`, with a median of `r median(leader_cult$number_leader_records, na.rm=T)`, a mean of `r signif(mean(leader_cult$number_leader_records, na.rm=T), 3)`, and a standard deviation of `r signif(sd(leader_cult$number_leader_records, na.rm=T), 3)`. The text records were generally short, ranging from `r textstats$min` to `r textstats$max` words, with a median word count of `r textstats$median`, a mean of `r textstats$mean`, and standard deviation of `r textstats$sd`. Of the `r nrow(all_data)` text records, `r final_record_count` had evidence for at least one dimension specific to the current study.
